Reduced growth due to belowground sink limitation is not fully explained by reduced photosynthesis

### 2016-12-15 manuscript/reviewer_response_tree_physiology.Rmd

\
Dear Professor Mott,

We are grateful for the opportunity to resubmit our manuscript. We thank the reviewers for providing insightful comments to improve our manuscript. We have responded in detail to each comment, which has improved the overall quality of the manuscript. Below, we have itemized our response to each comment from both referees.

### **Referee #1**

\
- *Reviewer: 1. The role of nutrient defiency may have been underestimated...The relative contribution should be determined...*

\
*Response*: Unfortunately, we do not have the ability with this data set to determine the relative contribution of nutrient deficiency on the down regulation of photosynthesis. We were able to identify not one, but two processes that likely contributed to photosynthetic inhibition.  Both of these processes, nutrient limitation and starch accumulation, are discussed at length in the discussion. We absolutely agree that determining the contribution of each of these processes will provide valuable insights on how sink limitation impacts the carbon balance of plants.  By providing convincing empirical evidence of these processes, this experimental desish should provide a framework for future work to quantify the contributions of each.

\
- *Reviewer: 2. Authors indicated that one of the principal question in the paper was to address which physiological process best explains the down-regulation of photosynthesis...the physiological processes involved have not been determined in the work*

\
*Response*: We agree with the reviewers point here. We have re-worded the first principle question to say "1) can we identify the main physiological processes that best explain the down-regulation of photosynthesis?" to more accurately address what we were able to test with the presented data.

\
- *Reviewer: 3. After reading the title, aim of the work and abstract, it seems that the manuscript focused specifically on carbon source-sink relations and regulation. In this regard, too much conclusions are derived from the presented data. Nevertheless, discussion and conclusions focused more accurately on the obtained results*

@@

\
*Response*: We have clarified in "Experimental design" that seedlings were nursery seed grown in 40cm tall tube-stock.

\
- *Reviewer: 2. For Free saplings, it is not clear if roots could have expanded beyond the limits of the subplot or not.*

\
*Response*: We have clarified in "Experimental design" that roots were able to grow unrestricted below border material.

\
- *Reviewer: 3 In relation to the “Photosynthetic parameters”, please, define Jmax and Vcmax.*

@@

\
*Response*: We have defined both *V*~cmax~ and *J*~max~ in the section "Photosynthetic parameters"

\
- *Reviewer: 4. In “Leaf, root and soil chemistry” indicate how plant samples were dried (temperature, drying time), how they were milled (type of mill used, model, etc.) and how soils were sieved (pore size) and milled (type of mill used, model, etc.)*

\
*Response*: We have clarified in “Leaf, root and soil chemistry” that samples were over dried at 60&deg;C to a constant mass and ball milled for analysis. We have also added the pore size, drying time and mill type to the explanation of soil processing.

\
- *Reviewer: 5. Clairify how did soil volume treatments relate to the 61 digitalized seedlings, which include different Eucalyptus species? Were the 61 digitized seedlings grown in different soil volumes*

\
*Response*: The digitized seedlings used here are a collection of digitized plants from differing environments including understory, glasshouse and plantation (see Duursma et al. 2012).  We selected 5 *Eucalpytus* species from this collection, which represent canopy structures of similar seedling types and sizes. Although the digitized plants are from separate experiments, they allowed us to estimate self shading of the *Eucalyptus* seedlings in this experiment. As this experiment had an ~4 month duration, leaf counts and total leaf area of seedlings were generally small. This resulted in only a modest adjustment of the self-shading factor to total canopy carbon uptake.

\
- *Reviewer: 6. Instead of using C concentration data (which they had), the authors assumed that C accounted for half of the final dry mass of seedlings. I think this is quite a strong assumption to make; differences in C concentration above 5% could be in place*

@@

\
*Response*: Unfortunately, we do not have available C concentration data for plant tissues. Our plant chemistry data set includes leaf and root tissue nitrogen content and leaf TNC and 13C. Consequently, we have to assume a C concentration for plant tissues. Our use of a 50% C content is likely very close to any assumption of C content that we have to make. For example, an average value of C content of angiosperm woody tissues from the DRYAD wood C database is ~48%.

\
- *Reviewer: 7. I would strongly recommend adopting a whole-plant approach and the consideration of all possible sources and sinks in mass-balance models. For example, root exudation can account for an important proportion of the total C budget of a plant, but yet this was not considered in the C budget calculations (nor even suggested as an important component to include in future models, see line 501)*

\
*Response*: The reviewer makes a very good point relating to how mass balance attempts should account for all possible fates of net C uptake. In this study, as in many, it is difficult to accurately measure every facet of C allocation throughout the plant and into the soil. As the reviewer point out, it is entirely possible that aspects of belowground sink limitation could have impacted patterns of root exudation in these seedlings. However, root exudation is just one of these possible fates, which also include woody tissue respiration, herbivory or VOC's. As we were not able measure all of these possible fates of net C uptake, we utilized the mass balance approach to estimate the residual difference between the modelled seedling C uptake and harvested C mass. It is for this reason that we report differences in CUE, as a more conservative approach in testing the coordination between photosynthesis and growth. Modelling the impacts of other potential fates of C, including root exudation, would require a priori knowledge of the effects of sink limitation/root restriction/container volume on each of these process. As so many of these processes are extremely difficult to measure, we do not yet have the ability to make such predictions. Possible impacts of root exudation are addressed in "Biomass partitioning under sink limitation", but we do not address it ubiquitously in the discussion because it is just one of many potential pathways of C loss.

\
- *Reviewer: 8. I think future empirical and C models should evaluate the inter-relationship between the different sinks and sources AT THE WHOLE PLANT level, and not just the leaf level as indicated in lines 499-501.*

@@

\
*Response*: We absolutely agree with the reviewer. We have made the editorial change to include the fate of C allocation at whole plant level, which includes the addition of root exudation from the previous comment.

\
- *Reviewer: 9. Lines 56-59. Yes, but beware the suggested approach also ignores part of the C budget of plants, because you are not accounting for root exudates.*

\
*Response*: see response to comment #7.  Our approach uses the residual C from mass balance to estimate all other potential C losses.

\
- *Reviewer: 10. Lines 71-72. Yes, indeed, C allocation belowground (both to roots and exudates) is enhanced under high CO2 supply.*

### 2016-12-15 manuscript/manuscript_pve.Rmd
final edits from reviewers, manusript and reply sent to coauthors

\
Forty-nine *Eucalyptus tereticornis* seedlings, 20 weeks old and approximately 40 cm tall in tube stock, were nursery seed grown from a single local Cumberland plain (western Sydney) cohort. Six additional seedlings were harvested before planting to measure initial leaf area and dry mass of leaves, stems and roots. Six container volumes were used ranging from 5 L to 35 L, with a 22.5 cm diameter, and lengths ranging from 15 to 100 cm. Containers were constructed of PVC pipe capped at the bottom end and were filled with local top soil. Soil in each container was packed to achieve a target soil bulk density that matched local soil conditions of 1.7 g cm^-3^. An Imidacloprid (BAYER CropScience) insecticide tablet was planted 5 cm below the roots of each seedling. Containers were planted flush with the soil surface inside metal sleeves, designed to minimize excess air space between the container and outside soil while also allowing for container removal. This allowed for soil temperatures in containers to reflect conditions of naturally planted (Free) seedlings. Each experimental block (n=7) contained a complete replicate set of six container volumes as well as one Free seedling, with 1 m^2^ spacing. For each Free seedling, used as the control, a 1 m^2^ subplot was excavated to the hard layer and replaced with the same soil used in each container. A border of root exclusion material was buried 0.25 m deep and extended 0.25 m above the ground surface around each subplot to exclude local vegetation, which was further kept out by periodic weeding. Roots of Free seedlings were allowed to grow unrestricted below the root exlusion border.

@@

\
Overall, SRL was higher in seedlings in containers compared to Free seedlings but only in some of the container size treatments (Table `r pvetab2`, *P* = 0.009). Fine root length density was significantly higher in the two smallest container sizes and was the lowest in the largest container size (Table `r pvetab2`, *P* < 0.001). Over the duration of the experiment SLA~f~ was higher in Free seedlings, but was not different across container sizes (Table `r pvetab1`, *P* < 0.001) and this pattern was evident beginning in the first gas exchange measurement campaign (*P* < 0.001).

##Leaf and root chemistry
Leaf N~f~ was significantly higher in Free seedlings and the largest container volume compared to the smaller container volumes at the onset of gas exchange measurements (6th week, *P* < 0.001). Throughout the remainder of the experiment the smallest container volume had significantly lower leaf N~f~ than other soil volumes, while Free seedlings maintained the highest leaf N~f~ (Table `r pvetab1`, *P* < 0.001). Leaf starch content in the smallest container was ca. double  that of Free seedlings (*P* = 0.039), while leaf soluble sugars did not differ across treatments throughout the experiment (Table `r pvetab1`). Differences in leaf starch between the Free seedling and the smallest container were evident during the first gas exchange campaign (*P* = 0.001). Root N was higher in Free seedlings compared to seedlings in containers but only for some of the container size treatments (Table `r pvetab2`).

@@

\
Neither $\Psi$~pd~ nor $\Psi$~l~ were different across treatments, with mean values of -0.27 and -1.2 MPa across all seedlings, respectively. Although *g*~s~ in Free seedlings was generally higher than those in containers (Table `r pvetab3`, *P* < 0.001), the mean rates for all seedlings were high at 0.37 mol m^-2^ s^-1^ and did not change throughout the course of the experiment. Additionally, leaf $\delta$^13^C at final harvest was not different across treatments (Table `r pvetab1`). Combined, these indices provide strong evidence that water stress was not apparent on these seedlings throughout the experiment. Soil N at harvest was not different across soil volumes (mean of .045 %), with minimal decreases from pre-planting values (mean of = .049 %). This similarity indicates that nutrient leaching from Free seedlings or from draining of containers following natural rainfall events did not differ between treatments.

##Whole-plant C balance
Modelled cumulative net leaf C gain over 120 days varied three-fold across containers sizes (23.1 - 79.9 g C). The Free seedling control had five-fold greater modelled total C gain than the average of container treatments (213.6 - 42.7&pm;8.2 g C). For seedlings in containers, an average of 67 &pm;0.01 % of modelled C gain was not allocated to observed biomass, compared to 59.2 % for the Free seedling control. Consequently, the estimated CUE of containerized seedlings was lower than the Free seedling control (Figure `r cuefig`). As a result, the observed reductions in leaf *A*~n~ with decreasing soil volume when integrated across the 120 day experiment were not large enough to explain the reduction in observed seedling biomass across the container size treatments.

@@

# Tables {.unnumbered}
**Table 1**.Responses of plant and leaf characteristics of *Eucalyptus tereticornis* seedlings to soil volume treatments. Each value reflects the mean (&pm; 1 standard error) for each treatment. Seedling mass and leaf $\delta$^13^C values are from final harvest. Values of SLA~f~, leaf starch, leaf sugars and leaf N~f~ represent overall means across measurement campaigns (n=6). Different letters represent significant differences between treatments. The volume effect *P* value represents the overall difference between seedlings with soil volume restriction and the control seedlings.

@@

dust(table1)%>%
  sprinkle_colnames("Volume (L)", 
                        "Seedling mass (g)", 
                        "SLA~f~ (m^2^ kg^-1^)",
                        "Leaf Starch (%)", 
                        "Leaf Sugars (%)", 
                        "Leaf N~f~ (%)",
                        paste0("Leaf ", "$\\delta$","^13^C"," (\u2030)")) %>%
  sprinkle(rows= 1, part="head",bold=TRUE)%>%
  sprinkle_print_method("markdown")
```

\
**Table 2**. Responses of root characteristics of *Eucalyptus tereticornis* seedlings to soil volume treatments. Each value reflects the mean (&pm; 1 standard error) for each treatment. All values are from the final harvest. Values for fine root length density (FRLD) were only calculated for seedlings in containers as Free seedlings had potentially unlimited soil volume to exploit. Different letters represent significant differences between treatments. The volume effect P value represents the overall difference between seedlings with soil volume restriction and the control seedlings, except for FRLD which represents only differences between seedlings in containers.
```{r table2, echo=FALSE, warning=FALSE}
dust(table2)%>%
  sprinkle_colnames("Volume (L)", 
                        "Root N (%)", 
                        "SRL (m g^-1^)",
                        "FRLD (m dm^-3^)") %>%
  sprinkle(rows= 1, part="head",bold=TRUE)%>%
  sprinkle_print_method("markdown")
```

>>>>

\
Forty-nine *Eucalyptus tereticornis* seedlings, 20 weeks old and approximately 40 cm tall in tube stock, were nursery seed grown from a single local Cumberland plain (western Sydney) cohort. Six additional seedlings were harvested before planting to measure initial leaf area and dry mass of leaves, stems and roots. Six container volumes were used ranging from 5 L to 35 L, with a 22.5 cm diameter, and lengths ranging from 15 to 100 cm. Containers were constructed of PVC pipe capped at the bottom end and were filled with local top soil. Soil in each container was packed to achieve a target soil bulk density that matched local soil conditions of 1.7 g cm^-3^. An Imidacloprid (BAYER CropScience) insecticide tablet was planted 5 cm below the roots of each seedling. Containers were planted flush with the soil surface inside metal sleeves, designed to minimize excess air space between the container and outside soil while also allowing for container removal. This allowed for soil temperatures in containers to reflect conditions of naturally planted (Free) seedlings. Each experimental block (n=7) contained a complete replicate set of six container volumes as well as one Free seedling, with 1 m^2^ spacing. For each Free seedling, used as the control, a 1 m^2^ subplot was excavated to the hard layer and replaced with the same soil used in each container. A border of root exclusion material was buried 0.25 m deep and extended 0.25 m above the ground surface around each subplot to exclude local vegetation, which was further kept out by periodic weeding. Roots of Free seedlings were allowed to grow unrestricted below the root exclusion border.

@@

\
Overall, SRL was higher in seedlings in containers compared to Free seedlings but only in some of the container size treatments (Table `r pvetab1`, *P* = 0.009). Fine root length density was significantly higher in the two smallest container sizes and was the lowest in the largest container size (Table `r pvetab1`, *P* < 0.001). Over the duration of the experiment SLA~f~ was higher in Free seedlings, but was not different across container sizes (Table `r pvetab1`, *P* < 0.001) and this pattern was evident beginning in the first gas exchange measurement campaign (*P* < 0.001).

##Leaf and root chemistry
Leaf N~f~ was significantly higher in Free seedlings and the largest container volume compared to the smaller container volumes at the onset of gas exchange measurements (6th week, *P* < 0.001). Throughout the remainder of the experiment the smallest container volume had significantly lower leaf N~f~ than other soil volumes, while Free seedlings maintained the highest leaf N~f~ (Table `r pvetab2`, *P* < 0.001). Leaf starch content in the smallest container was ca. double  that of Free seedlings (*P* = 0.039), while leaf soluble sugars did not differ across treatments throughout the experiment (Table `r pvetab2`). Differences in leaf starch between the Free seedling and the smallest container were evident during the first gas exchange campaign (*P* = 0.001). Root N was higher in Free seedlings compared to seedlings in containers but only for some of the container size treatments (Table `r pvetab2`).

@@

\
Neither $\Psi$~pd~ nor $\Psi$~l~ were different across treatments, with mean values of -0.27 and -1.2 MPa across all seedlings, respectively. Although *g*~s~ in Free seedlings was generally higher than those in containers (Table `r pvetab3`, *P* < 0.001), the mean rates for all seedlings were high at 0.37 mol m^-2^ s^-1^ and did not change throughout the course of the experiment. Additionally, leaf $\delta$^13^C at final harvest was not different across treatments (Table `r pvetab2`). Combined, these indices provide strong evidence that water stress was not apparent on these seedlings throughout the experiment. Soil N at harvest was not different across soil volumes (mean of .045 %), with minimal decreases from pre-planting values (mean of = .049 %). This similarity indicates that nutrient leaching from Free seedlings or from draining of containers following natural rainfall events did not differ between treatments. 
##Whole-plant C balance
Modelled cumulative net leaf C gain over 120 days varied three-fold across containers sizes (Table `r pvetab1`). The Free seedling control had five-fold greater modelled total C gain than the average of container treatments (213.6 - 42.7&pm;8.2 g C). For seedlings in containers, an average of 67 &pm;0.01 % of modelled C gain was not allocated to observed biomass, compared to 59.2 % for the Free seedling control. Consequently, the estimated CUE of containerized seedlings was lower than the Free seedling control (Figure `r cuefig`). As a result, the observed reductions in leaf *A*~n~ with decreasing soil volume when integrated across the 120 day experiment were not large enough to explain the reduction in observed seedling biomass across the container size treatments.

@@

# Tables {.unnumbered}
**Table 1**. Responses of plant characteristics of *Eucalyptus tereticornis* seedlings to soil volume treatments, including modelled cumulative net leaf carbon gain. Empirical values reflect the mean (&pm; 1 standard error) for each treatment. Seedling mass, SRL, and fine root length density (FRLD) values are from final harvest. Values of SLA~f~, leaf starch, leaf sugars and leaf N~f~ represent overall means across measurement campaigns (n=6). Values for FRLD were only calculated for seedlings in containers as Free seedlings had potentially unlimited soil volume to exploit. For empirical values, different letters represent significant differences between treatments. The volume effect *P* value represents the overall difference between seedlings with soil volume restriction and the control seedlings, except for FRLD which represents only differences between seedlings in containers.

@@

dust(table1)%>%
  sprinkle_colnames("Volume (L)", 
                    "Seedling mass (g)", 
                    "SLA~f~ (m^2^ kg^-1^)",
                    "SRL (m g^-1^)",
                    "FRLD (m dm^-3^)",
                    "Total Net C Gain (g)") %>%
  sprinkle(rows= 1, part="head",bold=TRUE)%>%
  sprinkle_print_method("markdown")
```

\
**Table 2**. Responses of leaf and root chemistry of *Eucalyptus tereticornis* seedlings to soil volume treatments. Each value reflects the mean (&pm; 1 standard error) for each treatment. Values of leaf starch, leaf sugars and leaf N~f~ represent overall means across measurement campaigns (n=6). Leaf $\delta$^13^C and root N values are from final harvest. Different letters represent significant differences between treatments. The volume effect P value represents the overall difference between seedlings with soil volume restriction and the control seedlings.
```{r table2, echo=FALSE, warning=FALSE}
dust(table2)%>%
  sprinkle_colnames("Volume (L)", 
                    "Leaf Starch (%)", 
                    "Leaf Sugars (%)", 
                    "Leaf N~f~ (%)",
                    paste0("Leaf ", "$\\delta$","^13^C"," (\u2030)"),
                    "Root N (%)") %>%
  sprinkle(rows= 1, part="head",bold=TRUE)%>%
  sprinkle_print_method("markdown")
```
